Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-05-11 pituba/article/article_spe.tex
hotel writings.

\begin{abstract}
This work derives the classical material balance equation from the more general black-oil model. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. It is shown that spatially averaged PVT properties are required for material balance approach in general, though the averaging process may be neglected for many practical purposes.  An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined. Finally, a scheme for numerical solution is proposed.

Material balance equation and black-oil models are approaches widely employed in reservoir engineering though few works explore their connection beyond the general intuitive notion. 
%
%This work derives the classical material balance equation from a black-oil model. We examine the existing link between both approaches in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow. This extension consists in multiplying absolute (or single-phase) permeability ($\overleftrightarrow{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):

\begin{equation}\label{eq: exdarcy}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

@@

In order to close the system of equations, a connection among different phase pressures must be provided. Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\vec{k},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\vec{k},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

@@

%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields. Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
%
%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
%\begin{equation}
%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
%\end{equation}
%saturation discontinuities will be present due to gravity segregation.
%
%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
%\begin{equation}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
%\end{equation}
%
%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
%Cite Bear and Herrera.

@@

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha g\nabla z
\end{equation}
and, consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

>>>>

\begin{abstract}
This work derives the classical material balance equation from a three dimensional black-oil model. Sufficient conditions for solution uniqueness of material balance equation are presented. It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach, although it is also shown that the averaging process may be neglected for many practical purposes.  An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined. Finally, a scheme for numerical solution is proposed.

Material balance equation and black-oil models are widely employed in reservoir engineering, but few works have explored their connection.
%
%This work derives the classical material balance equation from a black-oil model. We examine the existing link between both approaches in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow. This extension consists in multiplying absolute (or single-phase) permeability ($\vec{\vec{k}}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):

\begin{equation}\label{eq: exdarcy}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

@@

In order to close the system of equations, a connection among different phase pressures must be provided. Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\vec{\vec{k}},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\vec{\vec{k}},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

@@

%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields. Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
%
%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
%\begin{equation}
%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
%\end{equation}
%saturation discontinuities will be present due to gravity segregation.
%
%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
%\begin{equation}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{\vec{k}}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
%\end{equation}
%
%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{\vec{k}}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
%Cite Bear and Herrera.

@@

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha g\nabla z
\end{equation}
and, consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}
